A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-26 _tex/manuscript.tex
Built site for gh-pages

To our knowledge, existing approaches for sample size planning for GLMMs
have exclusively focused on planning for power. Power analysis methods
for multilevel models can be categorized into formula-based methods and
simulation-based methods
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022}). Formula-based methods rely on formulas to calculate
power directly while simulation-based methods rely on repeatedly
simulating data with a known true effect size and estimating power
empirically (i.e., what percentage of simulated datasets produces a
significant p-value). Available formula-based software packages for
multilevel models often do not include GLMMs or are limited to very
simple designs
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022};
\citeproc{ref-westfallStatisticalPowerOptimal2014}{Westfall et al.,
2014}), making it necessary to build data simulations tailored
specifically to the study design. A number of tutorials have been
published describing how to perform such simulation-based power analysis
for multilevel models
(\citeproc{ref-arendStatisticalPowerTwolevel2019}{Arend \& Schäfer,
2019}; \citeproc{ref-brysbaertPowerAnalysisEffect2018}{Brysbaert \&
Stevens, 2018};
\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}; \citeproc{ref-greenSIMRPackagePower2016a}{Green \& MacLeod,
2016}; \citeproc{ref-kainPracticalGuidePower2015}{Kain et al., 2015};
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021};
\citeproc{ref-lafitSelectionNumberParticipants2021}{Lafit et al., 2021};
\citeproc{ref-zimmerSampleSizePlanning2022}{Zimmer et al., 2022}).
However, many of these tutorials focus on linear mixed models (LMMs) and
the common study designs (but see
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021},
for a tutorial that also covers more advanced settings). This narrow
focus provides limited guidance for researchers faced with more complex
study designs, especially when little prior knowledge about plausible
effect sizes is available (see the discussion in
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}).
Simulation-based power analysis with GLMMs requires making a range of
assumptions: The (conditional) distribution assumption specifies the
distributional family for the outcome variable. Assumptions about the
random effects include the assumption of normality (i.e., that the
random effects follow a normal distribution) and the covariance
structure among the random effects (i.e., if and how they are
correlated). Making these decisions requires understanding the
underlying assumptions of the model and ensuring they align with the
characteristics of the data being analyzed. Existing tutorials often
rely on heuristics for specifying variance components (e.g., the
standard deviation of random intercepts) or assume that results from
meta-analyses or data from pilot studies are available to determine
plausible values for all model parameters. However, in practice,
knowledge about those parameters from prior studies is often limited,
which makes specifying assumptions a practical challenge (see the
discussion in (\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et
al., 2008}) and
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al.,
2021})).

INSERT TABLE 1 HERE!

In Table 1, we give a short review of existing R packages that can be
used for power analysis for GLMMs.

\section{When to use tailored data
simulation?}\label{when-to-use-tailored-data-simulation}

Performing tailored simulation-based sample size planning is more
complicated and time-consuming than using the existing software tools
outlined in TABLE 1. The most important circumstances under which
tailored simulation-based sample size planning is necessary are 1)
complex study designs, 2) complex statistical hypotheses, 3) planning
for precision, 4) no available prior studies or pilot data.

Requirements of real-world studies are often more complex than the
simplified designs assumed by many user-friendly software packages for
sample size planning. One frequent issue in applied data analysis is
missing data, and there can be various reasons for this
(\citeproc{ref-littleStatisticalAnalysisMissing2014}{Little \& Rubin,
2014}). For example, data can be missing completely at random (e.g.,
because an electronic measurement device randomly failed for some
technical reasons). Alternatively, subjects might systematically drop
out or produce missing data, but this dropout can be explained by some
attributes also measured in the dataset (e.g., older subjects have a
higher probability to refuse answering a question on income). In a more
complicated scenario, missing data in some variable is caused by the
measured attribute itself (e.g., wealthy people are more likely to
refuse reporting their income). Moreover, many experimental designs
contain conditions in which values of the predictor variables are
missing by design. This can make data analysis more complicated because
predictors have to be coded in specific ways that prevent the estimated
GLMM from becoming unidentified. Whether missing data has an effect on
the sample size planning depends on our theoretical assumptions on how
the missingness is caused. However, it is often challenging to decide
whether missing data can be safely ignored in the data analysis and
sample size planning process based on a merely theoretical approach
(\citeproc{ref-gomilaMissingDataExperiments2022}{Gomila \& Clark,
2022}). Tailored simulation-based approaches offer the possibility to
include the assumed process of how data become missing in the data
simulation, thereby determining the required sample based on simulated
datasets that contain missing values (for an example, see
\citeproc{ref-lanePowerStrugglesEstimating2018}{Lane \& Hennes, 2018}).
As a byproduct, the simulated datasets can also be used to test whether
the intended data analysis provides the expected (unbiased) results,
despite the missing data. Although GLMMs can handle a large variety of
outcome variables, researchers are becoming increasingly aware that many
datasets might profit from even more sophisticated models. Common
examples are zero-inflated outcomes, censoring, and nonlinear predictor
effects that can be modeled with the R packages glmmTMB
(\citeproc{ref-brooksGlmmTMBBalancesSpeed2017}{Brooks et al., 2017}) or
brms (\citeproc{ref-burknerAdvancedBayesianMultilevel2018}{Bürkner,
2018}). Tailored simulation-based approaches do not share the same
limitation than the existing software solutions for power analysis that
focus exclusively on GLMMs. As long as there is a software package
available to estimate the model of interest, it is always possible to
perform tailored simulation-based sample size planning.

The most common hypotheses tested in psychological research are of the
type \(H_0: \beta = 0\), where \(\beta\) is a slope or intercept of a
regression model. However, many research questions in psychology
actually require testing more complex statistical hypotheses. In the new
era of preregistration and registered reports
(\citeproc{ref-chambersPresentFutureRegistered2022}{Chambers \&
Tzavella, 2022}), most research questions should be tested with directed
hypotheses because good theories at least postulate whether some
psychological effect of interest is positive or negative. Even better
theories should be able to specify the smallest effect sizes of interest
(SESOI) that must be exceeded if the effect has any practical relevance
(\citeproc{ref-lakensEquivalenceTestingPsychological2018}{Lakens,
Scheel, et al., 2018}). In combination, this might require a test such
as \(H_0: \beta \leq 0.1\). More elaborate research questions often
require testing hypotheses that consist of a combination of model
parameters, for example testing simple slopes
(\citeproc{ref-preacherComputationalToolsProbing2006}{Preacher et al.,
2006}) with a hypothesis such as \(H_0: \beta_0 + \beta_1 \leq 0\). If
the research question consists only of a single hypothesis of this sort,
it might be possible to reduce the hypothesis to a single regression
coefficient by clever coding and/or centering of predictor variables.
However, interesting research questions often consist of combined
hypotheses that consist of more than one separate statistical hypothesis
(for a tutorial on contrast analysis in GLMMs, see
\citeproc{ref-schadHowCapitalizePriori2020}{Schad et al., 2020}). For
example, a combined null hypothesis \(H_0\) might consist of two single
null hypotheses \(H_{01}: \beta_1 \leq 0\) and
\(H_{02}: \beta_0 + \beta_1 \leq 0\). For some research questions, the
combined null hypothesis \(H_0\) would be rejected if both \(H_{01}\)
\emph{AND} \(H_{02}\) are rejected. For other research questions, the
combined null hypothesis \(H_0\) would be rejected if \(H_{01}\)
\emph{OR} \(H_{02}\) \emph{OR} both are rejected. If the global
hypothesis \(H_0\) is combined with \emph{OR}, the p-values of the
single hypotheses must be corrected for multiple testing to avoid
\(\alpha\)-inflation for the global hypothesis
(\citeproc{ref-dmitrienkoTraditionalMultiplicityAdjustment2013}{Dmitrienko
\& D'Agostino, 2013}). However, if the global hypothesis \(H_0\) is
combined with \emph{AND}, a correction for multiple testing is not
necessary but rather a mistake that unnecessarily reduces the power of
the global HT. None of the software packages for sample size planning in
table X can handle combined hypotheses as discussed here and only some
can handle directed hypotheses. In contrast, our case study will
demonstrate how we can test directed combined hypotheses with tailored
simulation-based sample size planning.

Although planning for precision
(\citeproc{ref-cummingNewStatisticsWhy2014}{Cumming, 2014};
\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a};
\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}) has
been increasingly discussed as a useful strategy for empirical research,
all available software packages for sample size planning with GLMMs are
based on power analysis. Therefore, researchers that want to want to
apply an estimation strategy in their studies, without testing any
statistical hypotheses, currently cannot use the software packages
outlined in TABLE 1. However, tailored simulation-based sample size
planning can easily handle the planning for precision approach
(\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}).
The only change in procedure is that instead of computing HTs for each
simulated dataset and estimating statistical power across repetitions,
CIs are computed for each simulated dataset, and the expected width is
estimated.

All frameworks for sample size planning require the user to make
assumptions about the expected effect size. Assuming the true effect is
of this size (or greater), one can compute the (minimum) power of a HT
or the (maximum) expected width of a CI. Existing software packages for
sample size planning for GLMMs usually require to provide the assumed
effect in the unit of some standardized measure of effect size. When the
researcher has access to similar studies or pilot data, providing such
standardized effect sizes is feasible. However, note that choosing
effect sizes based on small pilot studies is generally not recommended,
as those estimates can be heavily biased
(\citeproc{ref-albersWhenPowerAnalyses2018a}{Albers \& Lakens, 2018};
\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).
Providing an informed standardized effect size can be an almost
impossible challenge when no prior studies of pilot data are available.
This problem is further exacerbated by the fact that GLMMs are so
flexible that general heuristics of what should be considered a small
effect do not exist or are difficult to defend. In our experience, using
domain knowledge to construct a tailored data simulation is the only
solution to determine plausible effect sizes in the absence of prior
evidence. It would be possible to use these tailored simulations to
extract plausible values for standardized effect sizes that could then
be inserted in existing software packages for sample size planning.
However, we would argue that when tailored data simulations necessary to
determine effect sizes anyway, performing the whole sample size planning
in a customized way is preferred over using the existing software
packages.

@@

Although the details differ depending on the specific study
characteristics, each tailored simulation-based sample size planning
requires a series of steps and decisions. We will introduce each step in
a theoretical section, followed by the practical application based on a
hypothetical case study. All code in this manuscript and simulation
results are available in the project's repository on the Open Science
Framework (\url{https://osf.io/dhwf4/}).

@@

The first step in every research process should be a clear definition of
the theoretical \emph{estimand}
(\citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al., 2021}),
i.e.~the theoretical quantity which is necessary to answer a specific
research question. The estimand consists of a quantity, that can be
described for each unit under investigation and a clear definition of
the target population, for which the quantity is of interest. For
example, an estimand might be the probability that a clinical
psychologists makes the correct diagnosis for a psychiatric patient with
major depression, averaged across all clinical psychologists and
depressed patients in psychiatric institutions in Germany.

The estimand should always be defined outside of any statistical model,
because there are usually a range of statistical methods that can be
used to estimate the same estimand, depending on the study design (e.g.,
a randomized experiment) that will produce the observed data in the
planned study. For many common research questions in psychology, the
estimand can be expressed as a statistical quantity that can be
estimated with a regression model, for example a single \(\beta\)
coefficient. However, this is not possible for all estimands, which is
why the literature discusses many estimation strategies beyond
regression
(\citeproc{ref-deffnerCausalFrameworkCrossCultural2022}{Deffner et al.,
2022}; \citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al.,
2021}).

@@

In the present hypothetical case study, we consider the effectiveness of
feedback provided by an artificial intelligence (AI) embedded in a
diagnostic decision support system. The context is a clinical setting,
where expert radiologists and students under training must detect
bleeding based on head scans from computer tomography (CT). In the
investigated AI-enabled diagnostic decision support system, an AI model
can provide initial diagnostic advice, which can be used as guidance by
the humans who are required to make the final diagnostic decision. The
research goal is to validate the effectiveness of the AI-enabled advice.
We consider the AI-enabled advice as effective, if the following pattern
holds, which we will first describe verbally:

\emph{We expect that for BOTH expert radiologists and medical students,
correct AI-advice leads to a higher probability of accurately diagnosing
a CT scan compared to no AI-advice presented, AND, we expect that for
BOTH experts and non-experts, incorrect advice leads to a lower
probability of accurately diagnosing a CT scan compared to no advice
presented.}

@@

\[
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average expert, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, average expert, average scan})
\end{aligned}
\] \[
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, average student, average scan})
\end{aligned}
\]

\[
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, average student, average scan})
\end{aligned}
\] For example, the first expression is concerned with the difference
between the probability that a correct diagnosis is made if correct
AI-advice is presented and the probability that a correct diagnosis is
made if no AI-advice is presented. This contrast is quantified for a
hypothetical \emph{typical} expert and a \emph{typical} scan, where
\emph{typical} is usually defined as an average score on all attributes
of the expert or scan.\footnote{Note that a different estimand would be
  the so-called average treatment effect (ATE). For the ATE, the
  probability contrast is defined for each combination of expert and
  scan, and then these contrasts are averaged across all experts and
  scans from the target population
  (\citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al., 2021}).}.
To complete our definition of the estimand, we have to define our target
population that consists of persons, stimuli, and AI-advice: With
respect to persons, we are only interested in expert radiologists and
medical students at German universities. With respect to stimuli, we are
only interested in the head CT scans made from subjects that do or do
not suffer from intracerebral hemorrhage. Lastly, we are only interested
in AI-advice given by a specific AI-enabled diagnostic decision support
system.

>>>>

To our knowledge, existing approaches for sample size planning for GLMMs
have exclusively focused on planning for power. In
Table~\ref{tbl-rpackages}, we review available software packages that
can be used to perform power analysis for multilevel models in R
{[}Version 4.4.0; R Core Team (\citeproc{ref-R-base}{2024}){]}. Power
analysis methods can be categorized into formula-based,
summary-statistics-based and simulation-based methods
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022}). Formula-based methods rely on exact formulas to
calculate power directly. Summary-statistics-based methods use
statistical theory to approximate power based on formula-based methods
developed for simple t-tests. Simulation-based methods rely on
repeatedly simulating data with a known true effect size and estimating
power empirically, that is what percentage of simulated datasets
produces a significant p-value. Available formula-based and
summary-statistics-based software packages for multilevel models often
do not include GLMMs or are limited to simple designs
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022};
\citeproc{ref-westfallStatisticalPowerOptimal2014}{Westfall et al.,
2014}), making it necessary to build data simulations tailored
specifically to the study design. A number of tutorials have been
published describing how to perform such simulation-based power analysis
for multilevel models
(\citeproc{ref-arendStatisticalPowerTwolevel2019}{Arend \& Schäfer,
2019}; \citeproc{ref-brysbaertPowerAnalysisEffect2018}{Brysbaert \&
Stevens, 2018};
\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}; \citeproc{ref-greenSIMRPackagePower2016a}{Green \& MacLeod,
2016a}; \citeproc{ref-kainPracticalGuidePower2015}{Kain et al., 2015};
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021};
\citeproc{ref-lafitSelectionNumberParticipants2021}{Lafit et al., 2021};
\citeproc{ref-zimmerSampleSizePlanning2022}{Zimmer et al., 2022}).
However, many of these tutorials focus on LMMs and the most common study
designs (see \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et
al., 2021} for a tutorial that also covers more advanced settings). This
narrow focus provides limited guidance for researchers using more
complex study designs, especially when little prior knowledge about
plausible effect sizes is available (see the discussion in
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}).
Simulation-based power analysis with GLMMs requires making a range of
assumptions about the model structure that should align with the
characteristics of the data being analyzed. Existing tutorials often
rely on heuristics for specifying variance components (e.g., the
standard deviation of random intercepts) or assume that results from
meta-analyses or data from pilot studies are available to determine
plausible values for all model parameters. However, in practice,
knowledge about those parameters from prior studies is often limited,
which makes specifying assumptions a practical challenge (see the
discussion in Maxwell et al.
(\citeproc{ref-maxwellSampleSizePlanning2008}{2008}) and Kumle et al.
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{2021})). We will
discuss a number of strategies on how to specify model parameters for
application-specific, tailored data simulations in a later chapter.

{\caption{{Overview of software packages for power analysis for
(generalized) linear mixed models in R.}{\label{tbl-rpackages}}}
\vspace{-20pt}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1165}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.3947}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0526}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0489}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2030}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1842}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Package name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Summary
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
GUI available
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
GLMM support
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Design limitations
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
glmmrBase & Formula- and simulation-based power analysis; non-linear
fixed effects and flexible covariance functions & no & yes & none &
Watson (\citeproc{ref-watsonGeneralisedLinearMixed2023}{2023}) \\
longpower & Formula-based power analysis; focus on longitudinal data &
\href{https://atrihub.shinyapps.io/power/}{yes} & no & limited set of
study designs with two levels & Iddi and C Donohue
(\citeproc{ref-iddiPowerSampleSize2022}{2022}) \\
mixedpower & Simulation-based power analysis; based on `simr' package &
no & yes & none & Kumle et al.
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{2021}) \\
mlmpower & Simulation-based power analysis; missing data mechanisms & no
& no & limited to two levels & Enders et al.
(\citeproc{ref-endersSimpleMonteCarlo2023}{2023}) \\
pamm & Simulation-based power analysis; based on `lme4' package & no &
no & none & Martin et al.
(\citeproc{ref-martinMeasuringIndividualDifferences2011}{2011}) \\
pass.lme & Formula-based power analysis; limited documentation & no & no
& none & Marco Chak Yan YU
(\citeproc{ref-marcochakyanyuPassLmePower2019}{2019}) \\
PowerAnalysisIL & Simulation-based power analysis; focus on longitudinal
data &
\href{https://github.com/ginettelafit/PowerAnalysisIL?tab=readme-ov-file\#shiny-app-and-r-package-to-perform-power-analysis-to-select-the-number-of-participants-in-intensive-longitudinal-studies}{yes}
& no & limited set of study designs with two levels & Lafit et al.
(\citeproc{ref-lafitSelectionNumberParticipants2021}{2021}) \\
powerlmm & Simulation-based power analysis; missing data mechanisms;
focus on longitudinal data &
\href{https://github.com/rpsychologist/powerlmm?tab=readme-ov-file\#launch-interactive-web-application}{yes}
& no & limited set of study designs with two or three levels & Magnusson
et al.
(\citeproc{ref-magnussonConsequencesIgnoringTherapist2018}{2018}) \\
simglm & Simulation-based power analysis; missing data mechanisms &
\href{https://simglm.brandonlebeau.org/reference/run_shiny.html}{yes} &
yes & limited to two or three levels & LeBeau
(\citeproc{ref-lebeauSimglmSimulateModels2017}{2017}) \\
simr & Simulation-based power analysis; based on `lme4' package & no &
yes & none & Green and MacLeod
(\citeproc{ref-greenSIMRPackagePower2016}{2016b}) \\
summary\_statistics based\_power & Summary-statistics-based power
analysis; based on `pwr' package &
\href{https://koumurayama.shinyapps.io/summary_statistics_based_power/}{yes}
& yes & limited set of study designs with two levels & Murayama et al.
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{2022}) \\
WebPower & Formula-based power analysis; based on `longpower' package;
focus on longitudinal data &
\href{https://webpower.psychstat.org/wiki/start}{yes} & no & limited set
of study designs with two levels & Zhang and Yuan
(\citeproc{ref-zhangPracticalStatisticalPower2018}{2018}) \\
\end{longtable}

{\vspace{-20pt}
\noindent \emph{Note.} GUI available = whether a graphical user interface is provided (usually a Shiny app that is available online or can be downloaded to run locally), GLMM support = whether power analysis can be performed for generalized linear mixed model (otherwise only linear mixed models are supported). Design limitations = whether only a limited selection of study designs can be specified (otherwise arbitrary designs within the model class are supported).}

\section{Reasons to use tailored data
simulation}\label{reasons-to-use-tailored-data-simulation}

Performing tailored simulation-based sample size planning is more
complicated and time-consuming than using the existing software tools
outlined in Table~\ref{tbl-rpackages}. Circumstances under which
tailored simulation-based sample size planning is necessary include 1)
complex study designs, 2) complex statistical hypotheses, 3) planning
for precision, and 4) lack of prior studies or pilot data.

First, requirements of real-world studies are often more complex than
the simplified designs assumed by many user-friendly software packages
for sample size planning. One frequent issue in applied data analysis is
missing data (\citeproc{ref-littleStatisticalAnalysisMissing2014}{Little
\& Rubin, 2014}). Missings can be completely random (e.g., an electronic
measurement device randomly failed for technical reasons), depend on
some attributes also measured in the dataset (e.g., older subjects
refuse answering a question on income), or be caused by the measured
attribute itself (e.g., wealthy people refuse to report their income).
Moreover, many experimental designs contain conditions in which values
of the predictor variables are missing by design. This can make data
analysis more complicated because predictors have to be coded in
specific ways that prevent the estimated GLMM from becoming
unidentified. Whether missing data has an effect on the sample size
planning depends on our theoretical assumptions on how the missingness
is caused. However, it is often challenging to decide whether missing
data can be safely ignored in the data analysis and sample size planning
process based on a merely theoretical approach
(\citeproc{ref-gomilaMissingDataExperiments2022}{Gomila \& Clark,
2022}). Tailored simulation-based approaches offer the possibility to
include the assumed process of how data become missing in the data
simulation, thereby determining the required sample size based on
simulated datasets that contain missing values (for example, see
\citeproc{ref-lanePowerStrugglesEstimating2018}{Lane \& Hennes, 2018}).
As a byproduct, the simulated datasets can also be used to test whether
the intended data analysis provides the expected (unbiased) results,
despite the missing data. Although GLMMs can handle a large variety of
outcome variables, researchers are becoming increasingly aware that many
datasets might profit from even more sophisticated models. Common
examples are zero-inflated outcomes, censoring, and nonlinear predictor
effects that can be modeled with the R packages \emph{glmmTMB}
(\citeproc{ref-brooksGlmmTMBBalancesSpeed2017}{Brooks et al., 2017}) or
\emph{brms}
(\citeproc{ref-burknerAdvancedBayesianMultilevel2018}{Bürkner, 2018}).
Tailored simulation-based approaches do not share the same limitations
as the existing software solutions for power analysis that focus
exclusively on GLMMs. As long as there is a software package available
to estimate the model of interest, it is always possible to perform
tailored simulation-based sample size planning.

Second, the most common hypotheses tested in psychological research are
of the type \(H_0: \beta = 0\), where \(\beta\) is a slope or intercept
of a regression model. However, many research questions in psychology
actually require testing more complex statistical hypotheses. In the new
era of preregistration and registered reports
(\citeproc{ref-chambersPresentFutureRegistered2022}{Chambers \&
Tzavella, 2022}), most research questions should be tested with directed
hypotheses because good theories at least postulate whether some
psychological effect of interest is positive or negative. Even better
theories should be able to specify the smallest effect sizes of interest
(SESOI) that must be exceeded if the effect has any practical relevance
(\citeproc{ref-lakensEquivalenceTestingPsychological2018}{Lakens,
Scheel, et al., 2018}). This might require a test such as
\(H_0: \beta \leq 0.1\). More elaborate research questions often require
testing hypotheses that consist of a combination of model parameters,
for example, testing simple slopes
(\citeproc{ref-preacherComputationalToolsProbing2006}{Preacher et al.,
2006}) with a hypothesis such as \(H_0: \beta_0 + \beta_1 \leq 0\). If
the research question consists only of a single hypothesis of this sort,
it might be possible to reduce the hypothesis to a single regression
coefficient by clever coding and/or centering of predictor variables.
However, research questions often consist of combined hypotheses that
consist of more than one separate statistical hypothesis (for a tutorial
on contrast analysis in GLMMs, see
\citeproc{ref-schadHowCapitalizePriori2020}{Schad et al., 2020}). For
example, a combined null hypothesis \(H_0\) might consist of two single
null hypotheses \(H_{01}: \beta_1 \leq 0\) and
\(H_{02}: \beta_0 + \beta_1 \leq 0\). For some research questions, the
combined null hypothesis \(H_0\) would be rejected if both \(H_{01}\)
\emph{AND} \(H_{02}\) are rejected. For other research questions, the
combined null hypothesis \(H_0\) would be rejected if \(H_{01}\)
\emph{OR} \(H_{02}\) \emph{OR} both are rejected. If the global
hypothesis \(H_0\) is combined with \emph{OR}, the p-values of the
single hypotheses must be corrected for multiple testing to avoid
\(\alpha\)-inflation for the global hypothesis
(\citeproc{ref-dmitrienkoTraditionalMultiplicityAdjustment2013}{Dmitrienko
\& D'Agostino, 2013}). However, if the global hypothesis \(H_0\) is
combined with \emph{AND}, a correction for multiple testing is not
necessary but rather a mistake that unnecessarily reduces the power of
the global HT. None of the software packages for sample size planning in
Table~\ref{tbl-rpackages} can handle combined hypotheses and only some
can handle directed hypotheses. In contrast, our case study will
demonstrate how we can test directed combined hypotheses with tailored
simulation-based sample size planning.

Third, all available software packages for sample size planning with
GLMMs are based on power analysis and do not support planning for
precision. Thus, researchers that want to apply an estimation strategy
instead of testing statistical hypotheses
(\citeproc{ref-cummingNewStatisticsWhy2014}{Cumming, 2014}), currently
cannot use the software packages outlined in Table~\ref{tbl-rpackages}.
However, tailored simulation-based sample size planning can easily
handle the planning for precision approach
(\citeproc{ref-maxwellSampleSizePlanning2008}{Maxwell et al., 2008}).
The only change in procedure is that instead of computing HTs for each
simulated dataset and estimating statistical power across repetitions,
CIs are computed for each simulated dataset, and their expected width is
estimated.

Fourth, all frameworks for sample size planning require the user to make
assumptions about the expected effect size. Assuming the true effect is
of this size (or greater), one can compute the (minimum) power of a HT
or the (maximum) expected width of a CI. Existing software packages for
sample size planning for GLMMs usually require to provide the assumed
effect in the unit of some standardized measure of effect size. When the
researcher has access to similar studies or pilot data, providing such
standardized effect sizes is feasible. However, choosing effect sizes
based on small pilot studies is generally not recommended, as those
estimates can be heavily biased
(\citeproc{ref-albersWhenPowerAnalyses2018a}{Albers \& Lakens, 2018};
\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).
Providing an informed standardized effect size can be an almost
impossible challenge when no prior studies of pilot data are available.
This problem is further exacerbated by the fact that GLMMs are so
flexible that general heuristics of what should be considered a small
effect do not exist or are difficult to defend. In the absence of prior
evidence, using domain knowledge to construct a tailored data simulation
is considered the best solution to determine plausible effect sizes (see
scenario 3 in \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle
et al., 2021}). It would be possible to use these tailored simulations
to compute standardized effect sizes that could then be inserted in
existing software packages for sample size planning. However, we would
argue that when tailored data simulations are necessary to determine
effect sizes anyway, performing the whole sample size planning in a
customized way is preferred over using the existing software packages.

@@

Although the details differ for each study, every tailored
simulation-based sample size planning requires a series of steps and
decisions. We will introduce each step in a theoretical section,
followed by the practical application based on a case study. All code in
this manuscript and simulation results are available on the Open Science
Framework (\url{https://osf.io/dhwf4/}) and on our tutorial website
(\url{https://timo-ko.github.io/glmm_simulation_tutorial/}).

@@

The first step in every research process is a clear definition of the
theoretical \emph{estimand}
(\citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al., 2021}),
i.e.~the theoretical quantity which is necessary to answer a specific
research question. The estimand consists of a quantity that can be
described for each unit under investigation and a clear definition of
the target population, for which the quantity is of interest. For
example, an estimand might be the probability that a clinical
psychologist makes the correct diagnosis for a psychiatric patient with
major depression, averaged across all clinical psychologists and
depressed patients in psychiatric institutions in a given country.

The estimand should always be defined outside of any statistical model,
because there are usually a range of statistical methods that can be
used to estimate the same estimand, depending on the study design (e.g.,
a randomized experiment) that will produce the observed data in the
planned study. For many common research questions in psychology, it is
possible to express the estimand as a statistical quantity that can be
estimated with a regression model, for example a single \(\beta\)
coefficient. However, this is not possible for all estimands, which is
why the literature discusses many estimation strategies beyond
regression
(\citeproc{ref-deffnerCausalFrameworkCrossCultural2022}{Deffner et al.,
2022}; \citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al.,
2021}).

@@

In the present case study, we consider the effectiveness of feedback
provided by an artificial intelligence (AI) embedded in a clinical
decision support system. The context is a clinical setting, where task
experts (i.e., radiologists), and non-expert (i.e., medical students)
must detect bleedings based on head scans from computer tomography (CT).
The AI model can provide initial diagnostic advice, which can be used as
guidance by the humans who are required to make the final diagnostic
decision. The research goal is to validate the effectiveness of the
AI-enabled advice. We consider the AI-enabled advice effective, if the
following pattern holds:

\emph{We expect that for BOTH radiologists and medical students, correct
AI advice leads to a higher probability of accurately diagnosing a CT
scan compared to no AI advice presented, AND, we expect that for BOTH
task experts and non-experts, incorrect advice leads to a lower
probability of accurately diagnosing a CT scan compared to no advice
presented.}

@@

\[
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average expert, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{incorrect advice, average expert, average scan})
\end{aligned}
\] \[
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, average student, average scan})
\end{aligned}
\]

\[
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{incorrect advice, average student, average scan})
\end{aligned}
\] For example, the first expression is concerned with the difference
between the probability that a correct diagnosis is made if correct AI
advice is presented and the probability that a correct diagnosis is made
if no AI advice is presented. This contrast is quantified for a
hypothetical \emph{typical} expert and a \emph{typical} scan, where
\emph{typical} is usually defined as an average score on all attributes
of the expert or scan.\footnote{Note that a different estimand would be
  the so-called average treatment effect (ATE). For the ATE, the
  probability contrast is defined for each combination of expert and
  scan, and then these contrasts are averaged across all experts and
  scans from the target population
  (\citeproc{ref-lundbergWhatYourEstimand2021}{Lundberg et al., 2021}).}.
To complete our definition of the estimand, we have to define our target
population that consists of persons, stimuli, and AI advice: With
respect to persons, we are only interested in experts (i.e.,
radiologists) and non-experts (i.e., medical students). With respect to
stimuli, we are only interested in the head CT scans made from subjects
that do or do not suffer from intracerebral hemorrhage. Lastly, we are
only interested in AI advice given by a specific AI-enabled system.

Although the estimand is initially defined outside of a statistical
model, it is only useful if we find a way to estimate it based on
observed data. For our exemplary research question, it is possible to
construct an experimental study where all participants are confronted
with the same set of head CT scans, but the kind of AI advice given for
each scan is randomly assigned within participants. This random
intervention allows us to produce an empirical estimate of our estimand,
although, in reality, each person receives only one kind of AI advice
(correct advice, incorrect advice, no advice at all) for each scan. We
will see later how each of the probability expressions in our estimand
can be modeled with the same GLMM. Estimating this GLMM based on the
data observed in our planned study will produce an estimate for each
probability, and these estimates can then be combined to compute an
estimate for each of the four probability contrasts. For pedagogical
reasons, we will skip the concrete definition of our estimate until we
have discussed how to simulate data based on a concrete GLMM in the next
section.

\subsection{Step 2: Simulate the data-generating
process}\label{step-2-simulate-the-data-generating-process}
